---
title: 'Regression Analysis in Outcomes Research: A Practical Workshop'
subtitle: "**Student Follow-Along Version**"
author: "Outcomes Research Course"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: '3'

editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

# Load required packages
library(tidyverse)
library(broom)
library(knitr)
library(DiagrammeR)
library(gridExtra)
library(forcats)

# Set seed for reproducibility
set.seed(42)
```

## Learning Objectives

By the end of this workshop, you will be able to:

1.  Identify exposure (X) and outcome (Y) variables in research
    questions
2.  Understand the role of confounders in causal relationships
3.  Prepare and explore data for regression analysis
4.  Construct and interpret regression models in R
5.  Interpret both continuous and categorical predictors across linear,
    logistic, and Poisson regression
6.  Communicate statistical findings effectively
7.  Recognize common biases and limitations in regression analysis

------------------------------------------------------------------------

# Part 1: Conceptual Foundation

## 1.1 Understanding Relationships Between Variables

### The Research Question Framework

In outcomes research, we often ask: **"Does X affect Y?"**

-   **Y (Outcome):** The dependent variable we want to understand or
    predict
-   **X (Exposure/Predictor):** The independent variable(s) that may
    influence Y
-   **Confounders:** Variables that affect both X and Y, potentially
    distorting the relationship

### Example Research Question

> **Does physical activity (X) improve sleep quality (Y)?**

```{r concept-diagram, echo=FALSE, fig.height=4}
grViz("
digraph causal_diagram {
  graph [rankdir = LR, bgcolor = transparent]
  node [shape = box, style = filled, fillcolor = lightblue]

  PhysicalActivity [fillcolor = lightgreen, label = 'Physical Activity\n(X)']
  SleepQuality [fillcolor = salmon, label = 'Sleep Quality\n(Y)']
  Age [fillcolor = lightyellow]
  Stress [fillcolor = lightyellow]

  PhysicalActivity -> SleepQuality [label = '  Causal effect?', color = red, penwidth = 2]
  Age -> PhysicalActivity [label = ' ']
  Age -> SleepQuality [label = ' ']
  Stress -> PhysicalActivity [label = ' ']
  Stress -> SleepQuality [label = ' ']

  {rank = same; Age; Stress}
}
", height = 300)
```

### Key Concept: Confounding

A **confounder** is a variable that is associated with both the exposure
(X) and outcome (Y), but is NOT on the causal pathway between them.

Without adjusting for confounders, we might mistake correlation for
causation. For example, people with higher stress might exercise less
AND have poorer sleep quality, regardless of the true effect of physical
activity on sleep.

------------------------------------------------------------------------

## 1.2 Null and Alternative Hypotheses

Before analyzing data, we formulate **hypotheses** about the
relationship between X and Y.

### The Null Hypothesis (H₀)

The **null hypothesis** states there is **no effect** or **no
relationship**:

> H₀: Physical activity has no effect on sleep quality (β₁ = 0)

This is the "skeptic's position" — we assume no effect until the data
convinces us otherwise.

### The Alternative Hypothesis (H₁ or Hₐ)

The **alternative hypothesis** states there **is an effect**:

> H₁: Physical activity affects sleep quality (β₁ ≠ 0)

This can be:

-   **Two-sided:** Effect could be positive OR negative (β₁ ≠ 0)
-   **One-sided:** Effect is in a specific direction (β₁ < 0 or β₁ > 0)

### How Regression Tests Hypotheses

When we run a regression, R automatically tests:

| What we test | Null Hypothesis | What p-value tells us |
|------------------------|------------------------|------------------------|
| Each coefficient | β = 0 (no effect) | Probability of seeing this result if H₀ is true |

-   **p < 0.05:** Reject H₀ → Evidence of an effect (statistically
    significant)
-   **p ≥ 0.05:** Fail to reject H₀ → Insufficient evidence of an effect

### Important Caveats

-   **Statistical significance ≠ practical significance** — A 0.01 point
    increase in sleep quality might be statistically significant but not
    meaningful
-   **Failing to reject H₀ ≠ proving no effect** — It may mean we lack
    power to detect it
-   **p-values don't tell us effect size** — Always report confidence
    intervals

------------------------------------------------------------------------

## 1.3 The Regression Equation

Regression allows us to model the relationship between X and Y while
adjusting for confounders.

### The Regression Equation

**Simple regression:** $Y = \beta_0 + \beta_1X + \epsilon$

**Multiple regression:**
$Y = \beta_0 + \beta_1X_1 + \beta_2X_2 + ... + \beta_kX_k + \epsilon$

Multiple regression allows us to adjust for confounders by including
them in the model.

------------------------------------------------------------------------

# Part 2: Data Exploration & Preparation

## 2.1 Dataset Introduction

We will use the **Sleep Health and Lifestyle Dataset**, which comprises
400 rows and 13 columns covering a wide range of variables related to
sleep and daily habits.

**Key Features:**

-   **Comprehensive Sleep Metrics:** Sleep duration, quality of sleep,
    and sleep disorders
-   **Lifestyle Factors:** Physical activity levels, stress levels, and
    BMI categories
-   **Cardiovascular Health:** Blood pressure and heart rate
    measurements
-   **Demographics:** Sex, age, and occupation

**Research Question:** Does physical activity level affect quality of
sleep?

-   **Outcome (Y):** Quality of Sleep (1-10 scale)
-   **Exposure (X):** Physical Activity Level (minutes/day)
-   **Potential Confounders:** Age, sex, stress level, BMI category

## 2.2 Load and Prepare Dataset

```{r load-data}
# Load the dataset
data <- read_csv("Sleep_health_and_lifestyle_dataset (1).csv")

# Clean column names and prepare variables
# TODO: Rename columns to shorter, code-friendly names
# Hint: new_name = `Old Column Name`
data <- data %>%
  rename(
    person_id = `Person ID`,
    sleep_duration = ___,
    quality_of_sleep = ___,
    physical_activity = ___,
    stress_level = ___,
    bmi_category = `BMI Category`,
    blood_pressure = `Blood Pressure`,
    heart_rate = `Heart Rate`,
    daily_steps = `Daily Steps`,
    sleep_disorder = `Sleep Disorder`,
    # The dataset records biological sex (Male/Female), so we use Sex
    Sex = Gender
  ) %>%
  mutate(
    # Clean up BMI category: merge "Normal" and "Normal Weight"
    bmi_category = str_trim(bmi_category),
    bmi_category = case_when(
      bmi_category %in% c("Normal", "Normal Weight") ~ "Normal Weight",
      TRUE ~ bmi_category
    ),
    # TODO: Set reference categories for factor variables
    # Hint: fct_relevel(factor(variable), "Reference Level")
    Sex = fct_relevel(factor(Sex), ___),
    bmi_category = fct_relevel(factor(bmi_category), ___),
    sleep_disorder = fct_relevel(factor(sleep_disorder), ___),
    Occupation = factor(Occupation),
    # TODO: Create a binary high_activity variable (1 if >= 60 min, 0 otherwise)
    # Hint: ifelse(condition, value_if_true, value_if_false)
    high_activity = ifelse(___, 1, 0),
    high_activity_label = factor(
      ifelse(high_activity == 1, "High Activity", "Low Activity"),
      levels = c("Low Activity", "High Activity")
    ),
    age_group = cut(
      Age,
      breaks = c(0, 35, 45, 55, 100),
      labels = c("<35", "35-45", "46-55", ">55")
    )
  )

# Confirm the reference categories used in the workshop
levels(data$Sex)
levels(data$bmi_category)
levels(data$sleep_disorder)

# Preview the data
head(data) %>%
  select(person_id, Age, Sex, physical_activity, stress_level, quality_of_sleep)
```

### Why Reference Categories Matter

For categorical variables, regression coefficients are always interpreted
**relative to a reference category**. In this workshop, we explicitly set
the following reference categories:

-   `Sex`: **Female**
-   `bmi_category`: **Normal Weight**
-   `sleep_disorder`: **None**

This makes the model output easier to interpret than relying on the
default alphabetical ordering.

## 2.3 Exploratory Data Analysis

```{r summary-stats}
# TODO: Select the key numeric variables and display summary statistics
# Hint: select(var1, var2, ...) %>% summary()
data %>%
  select(___, ___, ___, ___, ___) %>%
  summary()
```

```{r exploratory-plots, fig.height=10}
# 1. Distribution of outcome (quality of sleep)
# TODO: Fill in the variable for the x-axis and choose the right geom
# Hint: geom_histogram() for distributions
p1 <- ggplot(data, aes(x = ___)) +
  geom_histogram(bins = 10, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Sleep Quality",
    x = "Quality of Sleep (1-10)",
    y = "Count"
  ) +
  theme_minimal()

# 2. Sleep quality by activity level (crude comparison)
# TODO: Fill in x, y, and fill aesthetics; choose the right geom
# Hint: geom_boxplot() for comparing groups
p2 <- ggplot(data, aes(
  x = ___,
  y = ___,
  fill = high_activity_label
)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Sleep Quality by Physical Activity Level (Crude Comparison)",
    x = "Activity Level",
    y = "Quality of Sleep (1-10)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Set2")

# 3. Relationship between physical activity and sleep quality
# TODO: Fill in x, y, and color aesthetics for a scatterplot
# Hint: geom_point() for scatterplots, geom_smooth(method = "lm") for trend lines
p3 <- ggplot(data, aes(x = ___, y = ___, color = ___)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Physical Activity vs. Sleep Quality (by Sex)",
    x = "Physical Activity (min/day)",
    y = "Quality of Sleep (1-10)",
    color = "Sex"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1")

# 4. Stress level distribution by activity level
p4 <- ggplot(data, aes(x = factor(stress_level), fill = high_activity_label)) +
  geom_bar(position = "fill") +
  labs(
    title = "Stress Level Distribution by Activity Level",
    x = "Stress Level (1-10)",
    y = "Proportion",
    fill = "Activity Level"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = scales::percent)

# Display all plots
grid.arrange(p1, p2, p3, p4, ncol = 2)
```

### Discussion Question

Look at the plots above. What do you observe about the relationship
between physical activity and sleep quality? Why might a simple
comparison be misleading?

------------------------------------------------------------------------

# Part 3: Building Regression Models

## 3.1 Simple Comparison (Crude Analysis)

First, let's look at the unadjusted difference in means.

```{r crude-comparison}
# Calculate mean difference
mean_diff <- data %>%
  group_by(high_activity_label) %>%
  summarise(mean_sleep_quality = mean(quality_of_sleep), .groups = "drop") %>%
  pivot_wider(
    names_from = high_activity_label,
    values_from = mean_sleep_quality
  ) %>%
  mutate(difference = `High Activity` - `Low Activity`)

print(mean_diff)
```

The crude analysis shows people with high physical activity have better
sleep quality. But is this the true effect or could there be
confounding?

------------------------------------------------------------------------

## 3.2 Simple Linear Regression

Now let's use regression to estimate the effect.

```{r simple-regression}
# TODO: Fit a simple linear regression predicting quality_of_sleep
#       from physical_activity
# Hint: lm(outcome ~ predictor, data = data)

model_simple <- lm(___ ~ ___, data = data)

# Display results
summary(model_simple)

# Tidy format
tidy_simple <- tidy(model_simple, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

print(tidy_simple)
```

**Key outputs:**

-   **Coefficient:** Estimated effect on sleep quality per minute of
    physical activity
-   **95% CI:** Range of plausible values
-   **p-value:** Statistical significance

### Sleep Quality by Sex

Let's also examine the relationship between sex and sleep quality.

```{r sex-sleep-quality}
# Compare sleep quality by sex
data %>%
  group_by(Sex) %>%
  summarise(
    n = n(),
    mean_quality = mean(quality_of_sleep),
    sd_quality = sd(quality_of_sleep),
    .groups = "drop"
  )

# TODO: Fit a simple regression with Sex as the predictor for quality_of_sleep
# Hint: lm(outcome ~ predictor, data = data)

model_sex <- lm(___ ~ ___, data = data)

summary(model_sex)
```

**Interpretation:** The coefficient for `SexMale` tells us the average
difference in sleep quality between males and females. Because Female is
the reference category, `SexMale` is interpreted as:

> average sleep quality among males minus average sleep quality among
> females

This idea will carry forward when we interpret categorical variables in
multiple linear, logistic, and Poisson regression.

------------------------------------------------------------------------

## 3.3 Multiple Linear Regression (Adjusting for Confounders)

Now let's adjust for confounders to get a better estimate of the effect.

```{r multiple-regression}
# TODO: Fit a multiple regression model predicting quality_of_sleep from:
#       physical_activity, Age, Sex, stress_level, and bmi_category
# Hint: lm(outcome ~ x1 + x2 + x3 + ..., data = data)

model_adjusted <- lm(
  ___ ~ ___ + ___ + ___ + ___ + ___,
  data = data
)

# Display results
summary(model_adjusted)

# Tidy format
tidy_adjusted <- tidy(model_adjusted, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

print(tidy_adjusted)
```

**Interpreting coefficients:**

-   **Continuous variables** (e.g., `physical_activity`, `Age`,
    `stress_level`):
    Each coefficient represents the expected change in sleep quality for
    a 1-unit increase in that predictor, **holding all other variables
    constant**.

-   **Binary categorical variables** (e.g., `Sex`):
    The coefficient compares the mean outcome for one category versus
    the reference category.

    -   `SexMale`: the average difference in sleep quality for males
        compared with females, adjusting for the other covariates

-   **Multi-level categorical variables** (e.g., `bmi_category`):
    Each coefficient compares one category to the chosen reference
    group.

    -   `bmi_categoryOverweight`: the average difference in sleep
        quality for overweight individuals compared with those of normal
        weight
    -   `bmi_categoryObese`: the average difference in sleep quality for
        individuals with obesity compared with those of normal weight

For example, if the coefficient for `SexMale` were -0.40, we would say:

> After adjusting for physical activity, age, stress level, and BMI
> category, males had an average sleep quality score that was 0.40
> points lower than females.

If the coefficient for `bmi_categoryObese` were -1.10, we would say:

> After adjustment, individuals with obesity had an average sleep
> quality score 1.10 points lower than individuals with normal weight.

------------------------------------------------------------------------

## 3.4 Interpreting Categorical Predictors Across Regression Models

Many predictors in outcomes research are **categorical**, such as sex,
BMI category, occupation, or disease status.

### Reference categories

When a categorical variable is included in a regression model, R creates
indicator variables and compares each category to a **reference
category**.

For example, if `Sex` has two categories:

-   **Female** = reference category
-   `SexMale` = comparison category

then the coefficient for `SexMale` tells us how males differ from
females, **holding all other variables constant**.

For variables with more than two categories, such as `bmi_category`, R
creates one coefficient for each non-reference level:

-   `bmi_categoryOverweight`: compares overweight to normal weight
-   `bmi_categoryObese`: compares obesity to normal weight

### General interpretation rule

-   **Linear regression:** coefficient = difference in the **mean
    outcome** compared with the reference group
-   **Logistic regression:** exp(coefficient) = **odds ratio** compared
    with the reference group
-   **Poisson regression:** exp(coefficient) = **rate ratio** or
    **incidence rate ratio (IRR)** compared with the reference group

### Template language

For a categorical predictor, you can usually write:

-   **Linear:** "Compared with [reference], those in [category] had an
    average outcome that was β units higher/lower."
-   **Logistic:** "Compared with [reference], those in [category] had OR
    times the odds of the outcome."
-   **Poisson:** "Compared with [reference], those in [category] had IRR
    times the expected count/rate of the outcome."

> **Important:** For a categorical variable with several levels (such as
> BMI category or occupation), each coefficient compares one level with
> the reference group. You can also test the variable **overall** to ask
> whether the categories differ as a whole.

------------------------------------------------------------------------

## 3.5 Comparing Models

```{r compare-models, fig.height=5}
# Compare coefficient estimates
comparison <- bind_rows(
  tidy_simple %>%
    filter(term == "physical_activity") %>%
    mutate(Model = "Unadjusted"),
  tidy_adjusted %>%
    filter(term == "physical_activity") %>%
    mutate(Model = "Adjusted")
) %>%
  select(Model, estimate, conf.low, conf.high, p.value)

print(comparison)

# Visualize the comparison
ggplot(comparison, aes(x = Model, y = estimate, color = Model)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.2,
    linewidth = 1
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Effect of Physical Activity on Sleep Quality",
    subtitle = "Comparing unadjusted vs. adjusted estimates",
    y = "Change in Sleep Quality (per min of activity)",
    x = ""
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
```

### Critical Insight

**Unadjusted:** Each additional minute of physical activity is
associated with a `r round(comparison$estimate[1], 3)` point increase in
sleep quality.

**Adjusted:** After controlling for age, sex, stress, and BMI, each
additional minute of physical activity is associated with a
`r round(comparison$estimate[2], 3)` point increase in sleep quality.

Adjusting for confounders helps isolate the effect of physical activity.

------------------------------------------------------------------------

## 3.6 Beyond Linear Regression: Logistic and Poisson

Linear regression works for **continuous outcomes**. For other outcome
types, we use different regression models:

| Outcome Type | Example | Regression Type | R Function |
|------------------|------------------|------------------|------------------|
| Continuous | Sleep quality score | Linear | `lm()` |
| Binary (Yes/No) | Sleep disorder | Logistic | `glm(..., family = binomial)` |
| Count | Daily steps (thousands) | Poisson | `glm(..., family = poisson)` |

### Logistic Regression (Binary Outcomes)

```{r logistic-example}
# Create a binary outcome: Does the person have a sleep disorder?
data <- data %>%
  mutate(has_disorder = ifelse(sleep_disorder != "None", 1, 0))

# TODO: Fit a logistic regression predicting has_disorder from:
#       physical_activity, Age, Sex, stress_level, and bmi_category
# Hint: glm(outcome ~ predictors, data = data, family = binomial)

model_logistic <- glm(
  ___ ~ ___ + ___ + ___ + ___ + ___,
  data = data,
  family = ___
)

# View results - coefficients are log-odds
tidy(model_logistic, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))
```

**Exponentiate to get Odds Ratios:**

```{r logistic-or}
# TODO: Use tidy() with exponentiate = TRUE to get Odds Ratios
# Hint: tidy(model, conf.int = TRUE, exponentiate = TRUE)

logistic_or <- tidy(model_logistic, conf.int = ___, exponentiate = ___) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  select(term, OR = estimate, conf.low, conf.high, p.value)

print(logistic_or)
```

**Interpreting categorical variables in logistic regression:**

After exponentiating the coefficients, categorical predictors are
interpreted as **odds ratios relative to the reference category**.

-   `SexMale`: the odds ratio comparing males with females
-   `bmi_categoryOverweight`: the odds ratio comparing overweight
    individuals with those of normal weight
-   `bmi_categoryObese`: the odds ratio comparing individuals with
    obesity with those of normal weight

For example:

-   If `SexMale` has OR = 1.40, males have **1.40 times the odds** of a
    sleep disorder compared with females, adjusting for the other
    variables.
-   If `bmi_categoryObese` has OR = 2.10, individuals with obesity have
    **2.10 times the odds** of a sleep disorder compared with those of
    normal weight.

An OR greater than 1 indicates **higher odds** than the reference group;
an OR less than 1 indicates **lower odds**.

**Continuous predictor reminder:** An OR of
`r round(exp(coef(model_logistic)["physical_activity"]), 2)` for
physical activity means that for each additional minute of daily
physical activity, the odds of having a sleep disorder change by a
factor of `r round(exp(coef(model_logistic)["physical_activity"]), 2)`,
adjusting for age, sex, stress level, and BMI category.

### Poisson Regression (Count Outcomes)

```{r poisson-example}
# Use daily steps as a count-like outcome (scaled to thousands)
data <- data %>%
  mutate(steps_thousands = round(daily_steps / 1000))

# TODO: Fit a Poisson regression predicting steps_thousands from:
#       physical_activity, Age, Sex, stress_level, and bmi_category
# Hint: glm(outcome ~ predictors, data = data, family = poisson)

model_poisson <- glm(
  ___ ~ ___ + ___ + ___ + ___ + ___,
  data = data,
  family = ___
)

# TODO: Exponentiate to get incidence rate ratios (IRR)
# Hint: tidy(model, conf.int = TRUE, exponentiate = TRUE)

poisson_irr <- tidy(model_poisson, conf.int = ___, exponentiate = ___) %>%
  mutate(across(where(is.numeric), ~ round(., 2))) %>%
  select(term, IRR = estimate, conf.low, conf.high, p.value)

print(poisson_irr)
```

**Interpreting categorical variables in Poisson regression:**

After exponentiating the coefficients, categorical predictors are
interpreted as **rate ratios** or **incidence rate ratios (IRRs)**
relative to the reference category.

-   `SexMale`: the expected step count for males relative to females
-   `bmi_categoryOverweight`: the expected step count for overweight
    individuals relative to those of normal weight
-   `bmi_categoryObese`: the expected step count for individuals with
    obesity relative to those of normal weight

For example:

-   If `SexMale` has IRR = 1.12, males have an expected daily step count
    that is **12% higher** than females, adjusting for the other
    covariates.
-   If `bmi_categoryObese` has IRR = 0.88, individuals with obesity have
    an expected daily step count that is **12% lower** than those of
    normal weight.

An IRR above 1 means a **higher expected count/rate** than the reference
group; an IRR below 1 means a **lower expected count/rate**.

**Continuous predictor reminder:** An IRR of
`r round(exp(coef(model_poisson)["physical_activity"]), 2)` for
physical activity means each additional minute of physical activity is
associated with a
`r round((exp(coef(model_poisson)["physical_activity"]) - 1) * 100, 1)`%
change in the expected daily step count.

### Quick Comparison: Interpreting Categorical Predictors Across Models

| Model type | Example coefficient | What it compares | Interpretation |
|------------------|------------------|------------------|------------------|
| Linear | `SexMale = -0.40` | Male vs Female | Males have an average sleep quality score 0.40 points lower than females |
| Logistic | `exp(SexMale) = 1.40` | Male vs Female | Males have 1.40 times the odds of sleep disorder compared with females |
| Poisson | `exp(SexMale) = 1.12` | Male vs Female | Males have an expected step count 12% higher than females |
| Linear | `bmi_categoryObese = -1.10` | Obese vs Normal Weight | Individuals with obesity have average sleep quality 1.10 points lower |
| Logistic | `exp(bmi_categoryObese) = 2.10` | Obese vs Normal Weight | Individuals with obesity have 2.10 times the odds of sleep disorder |
| Poisson | `exp(bmi_categoryObese) = 0.88` | Obese vs Normal Weight | Individuals with obesity have an expected step count 12% lower |

### Key Takeaway

-   **Linear:** coefficient = change in Y for 1-unit change in X; for
    categorical predictors, the coefficient is a difference in mean
    outcome vs the reference group
-   **Logistic:** exp(coefficient) = **Odds Ratio**; for categorical
    predictors, the OR compares each category with the reference group
-   **Poisson:** exp(coefficient) = **Rate Ratio / IRR**; for
    categorical predictors, the IRR compares each category with the
    reference group

> **Quick rule:** Always identify the **reference category first** before
> interpreting any categorical coefficient.

> **Note:** In practice, we also assess goodness of fit (R², AIC/BIC)
> and check model assumptions (linearity, normality of residuals,
> homoscedasticity, and appropriate mean-variance structure) before
> finalizing our conclusions.

------------------------------------------------------------------------

# Part 4: Communicating Results

## 4.1 Publication-Ready Tables

```{r publication-table}
# TODO: Create a professional results table from tidy_adjusted
# Steps:
#   1. Add a "95% CI" column by pasting conf.low and conf.high
#   2. Add a "Variable" column with readable names using case_when()
#   3. Select the columns: Variable, Coefficient, 95% CI, p-value

results_table <- tidy_adjusted %>%
  mutate(
    `95% CI` = paste0(
      "(",
      sprintf("%.3f", conf.low),
      " to ",
      sprintf("%.3f", conf.high),
      ")"
    ),
    Variable = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "physical_activity" ~ "Physical Activity (per min/day)",
      term == "Age" ~ "Age (per year)",
      term == "SexMale" ~ "Sex: Male (vs Female)",
      term == "stress_level" ~ "Stress Level (per unit)",
      term == "bmi_categoryOverweight" ~ "BMI: Overweight (vs Normal Weight)",
      term == "bmi_categoryObese" ~ "BMI: Obese (vs Normal Weight)",
      TRUE ~ term
    )
  ) %>%
  select(Variable, Coefficient = estimate, `95% CI`, `p-value` = p.value)

kable(
  results_table,
  caption = "Table 1. Multiple Linear Regression Results: Predictors of Sleep Quality",
  digits = 3
)
```

## 4.2 Writing Results

**Fill in the blanks** using the output from your adjusted model above:

"After adjusting for age, sex, stress level, and BMI category, physical
activity was associated with a **\_\_\_\_\_\_**
point increase in sleep quality per additional minute of daily activity
(95% CI: **\_\_\_\_\_\_** to **\_\_\_\_\_\_**,
p < **\_\_\_\_\_\_**)."

If you want to describe a categorical predictor, use the same logic but
reference the comparison group. For example:

"After adjustment, males had a sleep quality score of **\_\_\_\_\_\_**
points different from females (95% CI: **\_\_\_\_\_\_** to **\_\_\_\_\_\_**),
holding physical activity, age, stress level, and BMI category
constant."

------------------------------------------------------------------------

# Part 5: Hands-On Practice

## 5.1 Practice Exercise

**New Research Question:** Does stress level affect sleep duration?

Using the same Sleep Health and Lifestyle dataset, explore a different
relationship:

-   **Outcome (Y):** Sleep Duration (hours)
-   **Exposure (X):** Stress Level (1-10 scale)
-   **Potential Confounders:** Age, physical activity, occupation

```{r practice-data}
# Preview relevant variables for practice
head(data, 10) %>%
  select(person_id, Age, Occupation, sleep_duration, stress_level, physical_activity)
```

------------------------------------------------------------------------

## 5.2 Your Tasks

### Step 1: Explore the Data

```{r practice-step1, eval=FALSE}
# YOUR CODE HERE

# 1. Calculate summary statistics for sleep duration by stress level groups
#    Hint: Create stress categories (low: 1-3, medium: 4-6, high: 7-10)

# 2. Create visualizations:
#    - Distribution of sleep duration
#    - Sleep duration by stress level (scatterplot)
#    - Box plot of sleep duration by stress categories

# 3. Explore potential confounders:
#    How do Age, physical activity, and Occupation relate to stress?
```

**Hints:** Use `summary()`, `group_by()`, `ggplot()`, and think about
whether any categorical predictors (such as `Occupation`) need a
reference group for interpretation.

------------------------------------------------------------------------

### Step 2: Fit Regression Models

```{r practice-step2, eval=FALSE}
# YOUR CODE HERE

# 1. Fit a simple regression: sleep_duration ~ stress_level

# 2. Fit a multiple regression adjusting for potential confounders
#    Consider including: Age, physical_activity, Occupation

# 3. Compare the stress_level coefficients between models

# 4. If you include Occupation, identify the reference category and
#    explain how you would interpret one Occupation coefficient
```

**Hints:** Use `lm()` and `tidy()`. If you include a categorical
variable, remember that each coefficient compares one level with the
reference category.

------------------------------------------------------------------------

### Step 3: Interpret Results

```{r practice-step3, eval=FALSE}
# YOUR CODE HERE

# 1. Create a comparison table of unadjusted vs adjusted estimates for stress_level

# 2. Visualize the effect estimates with confidence intervals

# 3. Write a one-sentence interpretation of your findings

# 4. Optional:
#    Write one sentence interpreting an Occupation coefficient
```

**Answer:** What's the crude vs. adjusted effect of stress on sleep
duration? Did confounding matter? What are the limitations?

------------------------------------------------------------------------

## 5.3 Solution (Reveal After Attempting)

<details>

<summary>**Click to reveal solution**</summary>

```{r practice-solution, class.source = 'fold-show'}
# STEP 1: Explore the data

# Create stress categories
data <- data %>%
  mutate(stress_category = case_when(
    stress_level <= 3 ~ "Low (1-3)",
    stress_level <= 6 ~ "Medium (4-6)",
    TRUE ~ "High (7-10)"
  ))

# Set a reference category for Occupation if desired
data <- data %>%
  mutate(Occupation = fct_relevel(Occupation, levels(Occupation)[1]))

# Summary statistics
data %>%
  group_by(stress_category) %>%
  summarise(
    n = n(),
    mean_sleep = mean(sleep_duration),
    sd_sleep = sd(sleep_duration),
    mean_age = mean(Age),
    mean_activity = mean(physical_activity),
    .groups = "drop"
  )

# Visualizations
p1 <- ggplot(data, aes(x = sleep_duration)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Sleep Duration",
    x = "Sleep Duration (hours)",
    y = "Count"
  ) +
  theme_minimal()

p2 <- ggplot(data, aes(x = stress_level, y = sleep_duration)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "red") +
  labs(
    title = "Sleep Duration vs Stress Level",
    x = "Stress Level",
    y = "Sleep Duration (hours)"
  ) +
  theme_minimal()

grid.arrange(p1, p2, ncol = 2)

# STEP 2: Fit regression models

# Simple model
model_practice_simple <- lm(sleep_duration ~ stress_level, data = data)
tidy(model_practice_simple, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

# Adjusted model with a categorical covariate
model_practice_adjusted <- lm(
  sleep_duration ~ stress_level + Age + physical_activity + Occupation,
  data = data
)
tidy(model_practice_adjusted, conf.int = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(., 3)))

# STEP 3: Compare and visualize

comparison_practice <- bind_rows(
  tidy(model_practice_simple, conf.int = TRUE) %>%
    filter(term == "stress_level") %>%
    mutate(Model = "Unadjusted"),
  tidy(model_practice_adjusted, conf.int = TRUE) %>%
    filter(term == "stress_level") %>%
    mutate(Model = "Adjusted")
) %>%
  select(Model, estimate, conf.low, conf.high, p.value)

print(comparison_practice)

# Forest plot
ggplot(comparison_practice, aes(x = Model, y = estimate, color = Model)) +
  geom_point(size = 4) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.2,
    linewidth = 1
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    title = "Effect of Stress Level on Sleep Duration",
    y = "Change in Sleep Duration (hours per 1-unit increase in stress)"
  ) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "none")
```

**Interpretation:** After adjusting for age, physical activity, and
occupation, each 1-point increase in stress level was associated with a
`r round(tidy(model_practice_adjusted, conf.int = TRUE) %>% filter(term == "stress_level") %>% pull(estimate), 2)`
hour change in sleep duration (95% CI:
`r round(tidy(model_practice_adjusted, conf.int = TRUE) %>% filter(term == "stress_level") %>% pull(conf.low), 2)`
to
`r round(tidy(model_practice_adjusted, conf.int = TRUE) %>% filter(term == "stress_level") %>% pull(conf.high), 2)`).

If you included a categorical predictor such as `Occupation`, each
Occupation coefficient would be interpreted as the mean difference in
sleep duration compared with the chosen reference occupation, holding
the other variables constant.

</details>

------------------------------------------------------------------------

# Part 6: Key Takeaways & Resources

## 6.1 Key Takeaways

**Concepts:**

-   Identify X (exposure), Y (outcome), and confounders
-   Understand how confounding distorts relationships
-   Use regression to adjust for confounders
-   Interpret continuous predictors as changes per unit increase
-   Interpret categorical predictors relative to a reference group

**Skills:**

-   Explore data and visualize relationships
-   Fit and interpret regression models in R
-   Compare crude vs. adjusted estimates
-   Interpret categorical variables in linear, logistic, and Poisson
    models
-   Create tables and communicate findings

------------------------------------------------------------------------

## 6.2 Common Pitfalls to Avoid

```{r pitfalls, echo=FALSE}
pitfalls <- tibble(
  Pitfall = c(
    "Assuming correlation = causation",
    "Not checking for confounding",
    "Ignoring reference categories",
    "P-hacking",
    "Not considering unmeasured confounding"
  ),
  Description = c(
    "An observed association doesn't prove causation",
    "Failing to identify and adjust for confounders",
    "Misinterpreting categorical variable coefficients",
    "Running many models and only reporting significant results",
    "Assuming all confounders are measured and included"
  ),
  `How to Avoid` = c(
    "Be cautious in causal language; consider study design",
    "Always note which category is the reference (e.g., Female, Normal Weight)",
    "Pre-specify your analysis plan; report all models tested",
    "Acknowledge limitations in your discussion"
  )
)

print(pitfalls)
```

------------------------------------------------------------------------

# Appendix: Quick R Reference

```{r cheatsheet, eval=FALSE}
# Fit regression
model <- lm(y ~ x1 + x2, data = data)

# View results
summary(model)                    # Full output
tidy(model, conf.int = TRUE)      # Clean table (broom)
glance(model)                     # Model fit stats

# Diagnostics
plot(model)                       # Check assumptions

# Set a reference category for a factor
data$group <- forcats::fct_relevel(data$group, "Reference Level")

# Logistic or Poisson: exponentiate for OR / IRR
tidy(model, conf.int = TRUE, exponentiate = TRUE)
```

------------------------------------------------------------------------

## Session Info

```{r session-info}
sessionInfo()
```
