---
title: "PHAR 305 Final Project"
author: "Your Name Here"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

# Load required packages
library(tidyverse)
library(broom)
library(knitr)
```

------------------------------------------------------------------------

# Instructions

## Rename this file

Replace “lastname_firstname” in the file name with your own last and first name before beginning the project (e.g., Doe_Jane_PHAR305_finalproject.Rmd). The same naming convention will automatically apply when you knit this file to HTML.

## Overview

This project will assess your understanding of how to formulate and test
a scientific research question, conduct and interpret a regression
analysis, and contextualize your research findings. You will apply these
concepts to a dataset of your choice.

**Total Points: 60**

Deadline to submit knit HTML and RMD files on Canvas: April 7, 2026 at
11:59 PM

## Requirements

1.  **Choose ONE dataset** from the options provided on Canvas
2.  Complete ALL sections of this analysis template
3.  Show all code and provide written interpretations
4.  Knit your final document to HTML before submitting

## Grading Criteria

| Section | Topic                          | Points |
|---------|--------------------------------|--------|
| Part 0  | Dataset Identified             | 1      |
| Part 1  | Research Question & Hypotheses | 11     |
| Part 2  | Data Exploration               | 13     |
| Part 3  | Regression Analysis            | 15     |
| Part 4  | Communicating Results          | 6      |
| Part 5  | Limitations & Interpretation   | 6      |
| Overall | RMD Rendered Successfully      | 3      |
| Overall | Code readability               | 5      |
| Total   |                                | 60     |

------------------------------------------------------------------------

# Your Dataset Selection (/1 point)

**Provide the name of your chosen dataset**

*Your answer:*

------------------------------------------------------------------------

# Part 1: Research Question & Hypotheses (/11 points)

## 1.1 Load Your Dataset (1 point)

Load your chosen dataset.

```{r load-data}



```

## 1.2 Define Your Research Question (8 points)

Based on your chosen dataset, formulate a research question using the
PICO framework (5 points):

**Population**

*Your answer:*

**Intervention**

*Your answer:*

**Comparator**

*Your answer:*

**Outcome**

*Your answer:*

**Your Research Question (1 sentence)**

*Your answer:*

Now, identify the variables **in your dataset** that will allow you to
test this research question (3 points):

**Outcome (Y)** *What is your dependent variable?*

*Your answer:*

**Exposure (X)** *What is your main predictor of interest?*

*Your answer:*

**List 2-3 potential confounders for the X-Y relationship**

*Your answer:*

## 1.3 Justify Your Confounders (2 points)

For each confounder you identified, provide a 2-3 sentence conceptual
justification based on the definition of a confounder that we learned in
class. Include at least one reference to a published study to support
your justification for each confounder.

*Tip:* You can cite studies within markdown by including a URL, for
example: <https://pubmed.ncbi.nlm.nih.gov/>

**Confounder justification**

*Your answer:*

------------------------------------------------------------------------

# Part 2: Data Exploration (/13 points)

## 2.1 Summary Statistics (4 points)

Provide at least one informative summary statistic for each of your key
variables (outcome, exposure, and each of your confounders identified in
Section 1.2).

```{r summary-stats}
# YOUR CODE HERE
```

## 2.2 Distribution of Outcome Variable (3 points)

Create a visualization showing the distribution of your outcome
variable.

```{r outcome-distribution}
# YOUR CODE HERE
```

**Interpretation**

Describe the distribution of your outcome variable (e.g., shape, central
tendency, spread) in 2-3 sentences.

*Your answer:*

------------------------------------------------------------------------

## 2.3 Exposure-Outcome Relationship (3 points)

Create a visualization showing the relationship between your exposure
and outcome.

```{r exposure-outcome-plot}
# YOUR CODE HERE
```

**Interpretation**

Describe the crude (unadjusted) relationship you observe between
exposure and outcome in 3-4 sentences.

*Your answer:*

------------------------------------------------------------------------

## 2.4 Exploring Potential Confounding (3 points)

Show evidence that at least one of your confounders is associated with
BOTH the exposure and the outcome.

```{r confounder-exploration}
# Show how a confounder relates to both X and Y
# You can create plots or summary tables

# YOUR CODE HERE
```

**Interpretation**

Explain how this evidence supports that this variable could be a
confounder in 3-5 sentences.

*Your answer:*

------------------------------------------------------------------------

# Part 3: Regression Analysis (/15 points)

## 3.1 Choose the Appropriate Regression Model (2 points)

Based on your outcome variable, select the appropriate regression model:

**a) What type of outcome variable do you have? (1 sentence)**

*Your answer:*

**b) Which regression model will you use and why? (2-3 sentences)**

*Your answer:*

## 3.2 Unadjusted (Simple) Regression (4 points)

Fit a simple regression model with only your exposure variable as the
predictor. Use the appropriate model type based on your answer above.

```{r simple-regression}
# YOUR CODE HERE
```

**Interpretation of Exposure Effect**

Identify and interpret the appropriate coefficient in the context of
your research question (2-3 sentences).

*Your answer:*

Is this association statistically significant? Interpret the confidence
interval AND p-value (1-2 sentences).

*Your answer:*

## 3.3 Adjusted (Multiple) Regression (4 points)

Fit a multiple regression model adjusting for your confounders selected
in Section 1.2. Use the same model type as above.

```{r multiple-regression}
# YOUR CODE HERE
```

**Interpretation of Exposure Effect after adjusting for confounders**

Identify and interpret the appropriate coefficient in the context of
your research question (2-3 sentences).

*Your answer:*

Is this association statistically significant? Interpret the confidence
interval AND p-value (1-2 sentences).

*Your answer:*

## 3.4 Compare Unadjusted vs. Adjusted Estimates (5 points)

Create a numeric AND visual comparison of your unadjusted and adjusted
estimates for the exposure variable.

```{r compare-models}
# YOUR CODE HERE
```

```{r compare-plot}
# YOUR CODE HERE
```

**Analysis of Confounding**

How did the inclusion of confounders change the coefficient for your
exposure after adjustment? In your answer, describe the direction the
coefficient changed (toward or away from the null effect), and whether
your conclusion about statistical significance was affected (4-5
sentences).

*Your answer:*

------------------------------------------------------------------------

# Part 4: Communicating Results (/6 points)

## 4.1 Publication-Ready Table (2 points)

Create a professional results table for your adjusted regression model.

```{r publication-table}
# Create a formatted results table
# Include: Variable names, Coefficients, 95% CI, p-values

# YOUR CODE HERE
```

## 4.2 Written Results Statement (4 points)

Write a 5-7 sentence results paragraph suitable for a research paper.
Your statement should include:

-   The research question addressed
-   The adjusted effect estimate with units
-   The 95% confidence interval
-   Whether the effect is statistically significant
-   The confounders that were controlled for and how they impacted the
    effect estimate

**Results**

*Your answer:*

------------------------------------------------------------------------

# Part 5: Limitations & Interpretation (/6 points)

## 5.1 Limitations (3 points)

Discuss THREE limitations of your dataset or analysis. These could
include unmeasured confounding variables, measurement bias in data
collection, selection bias in study design, lack of external validity,
limitations in the statistical analysis, etc.

**Limitation 1**

Describe and explain the potential impact on results in 2-3 sentences.

*Your answer:*

**Limitation 2**

Describe and explain the potential impact on results in 2-3 sentences.

*Your answer:*

**Limitation 3**

Describe and explain the potential impact on results in 2-3 sentences.

*Your answer:*

## 5.2 Interpretation (3 points)

What is your conclusion on whether your exposure causes the outcome?
Provide an informative answer based on your response to Sections 4.2 and
5.1; do not simply state that correlation does not equal causation.
Incorporate your assessment of the strengths and weaknesses of your
dataset and/or analysis. Answer in 4-5 sentences.

**Interpretation**

*Your answer:*

------------------------------------------------------------------------

# Submission Checklist

Before submitting, verify that you have the following, successfully
rendered in HTML format:

-   [ ] Selected and loaded ONE dataset
-   [ ] Stated a clear research question with PICO, X, Y, and
    confounders
-   [ ] Justified why your confounders could affect both X and Y
-   [ ] Created summary statistics and visualizations
-   [ ] Fit an unadjusted (simple) regression model
-   [ ] Fit an adjusted (multiple) regression model
-   [ ] Compared unadjusted vs. adjusted estimates
-   [ ] Created a publication-ready table
-   [ ] Written a results statement
-   [ ] Discussed limitations and interpretation
-   [ ] Replaced "Your Name Here" with your actual name
-   [ ] Successfully knitted the document to HTML

------------------------------------------------------------------------

## Session Info

```{r session-info}
sessionInfo()
```
